Magic-free Code Extraction of Generic Programs and Recursion Schemes in Coq

### 2024-03-18 paper/paper.tex
Examples

Both mergesort and quicksort are divide-and-conquer algorithms that can be
captured by the structure of an hylomorphisms. The structure of the recursion
is that of a binary tree.  For example, in the case of quicksort, a list is split
into a pivot, the label of the node, and two sublists. We define the data
functor of trees as follows:
\begin{coqcode}
Inductive ITreeF L N X := i_leaf (l : L) | i_node (n : N) (l r : X)
\end{coqcode}
We define the functor as a container using the following shapes and positions:
\begin{coqcode}
Inductive Tshape L A := | Leaf (ELEM : L) | Node (ELEM : A).
Inductive Tpos := | Lbranch | Rbranch. 
\end{coqcode}
We define the tree container, \coq{TreeF}, in a straightforward way, by making
the positions of type \coq{Tpos} only valid in \coq{Node}.
We define a series of definitions for tree container constructors and 
destructors:
\mpav{The following algebra should be called a\_in or l\_in whatever}
\begin{coqcode}
Definition a_out {L A X : Type} : App (TreeF L A) X ~> ITreeF L A X.

@@

Definition c_split : Coalg (TreeF unit int) (list int).
|{ x ~> match x with
        | nil => a_leaf tt
        | cons h t => let (l, r) := List.partition (fun x => x <=? h) t in
                      a_node h l r
        end}|.
Defined.
\end{coqcode}
We prove that the coalgebra \coq{c_split} is recursive by showing
that it respects the ``less-than'' relation on the length of the lists.
We then define, and extract, the hylomorphism \coq{hylo merge r_split}, which
extracts to the following OCaml code:
\begin{ocamlcode}
let rec qsort = function
| [] -> [] | h :: t ->
  let (l, r) = partition (fun x0 -> leb x0 h) t in
  let x0 = fun e -> qsort (match e with | Lbranch -> l | Rbranch -> r) in
  app (x0 Lbranch) (h :: (x0 Rbranch))
\end{ocamlcode}
Note that Coq's code extraction is unable to inline \coq{x0}, but the resulting
code is similar to a hand-written \coq{qsort}.

The mergesort algorithm can be defined analogously and can be found in the formalisation. [REF LINK].

@@

% Putting everything together, Mergesort is defined as the following
% hylomorphism:
% \begin{coqcode}
% Definition msort := hylo merge tsplit.
% \end{coqcode}
% Note, however, that we could have composed the catamorphism and anamorphism,
% and used the deforestation result to optimise the intermediate binary tree
% away:
% \begin{coqcode}
% Definition msort : Ext (cata merge \o rana tsplit).
% calculate. rewrite (deforest l_in_out). simpl. reflexivity. Defined.
% \end{coqcode}
% In both cases, the result is the same, and it extracts to the usual divide and
% conquer definition of Mergesort in OCaml:
% \begin{ocamlcode}
% let rec msort x = match x with
% | [] -> x | _ :: l -> (match l with | [] -> x | _ :: _ ->
%      let (l0, r) = splitL x [] [] in
%      let cx = fun p -> match p with | Lbranch -> l0 | Rbranch -> r in
%      mergeL (msort (cx Lbranch)) (msort (cx Rbranch)))
% \end{ocamlcode}
\subsubsection{Fusing a divide-and-conquer computation}
\mvol{This is good, but it needs some text to connect it to the parent sub-section,
  like: ``As an aside, we can show how aonther traversal can be fused into the divide-and-conquer
  algorithm.''}
Suppose that we map a function to the result of sorting the list. We can
use our framework to fuse both computations. In particular, consider
the following definition:
\begin{coqcode}
Definition qsort_times_two : Ext (Lmap times_two \o hylo merge tsplit).
  calculate.  apply hylo_fusion_l. 
  (* calculate the fused stage *)
Defined.
\end{coqcode}
Here, \coq{Lmap times_two} is a list map function defined as a hylomorphism,
and \coq{times_two} multiplies every element of the list by two.
After applying hylo fusion and the necessary rewrites, the hylomorphism that we
extract is 
\coq{hylo (merge \o natural times_two) tsplit}. Here, \coq{natural} defines
a natural transformation between container extensions by applying a function
to the shapes that preserves the validity of the positions. In this case,
the function is \coq{times_two} that multiplies every pivot in the Quicksort
tree by two. The extracted OCaml code is a single recursive traversal:
\begin{ocamlcode}
let rec qsort_times_two = function
| [] -> []
| h :: t ->
  let (l, r) = partition (fun x0 -> leb x0 h) t in
  let x0 = fun p -> qsort_times_two (match p with
                                     | Lbranch -> l
                                     | Rbranch -> r) in
  app (x0 Lbranch) ((mul (Uint63.of_int (2)) h) :: (x0 Rbranch))
\end{ocamlcode}

\subsection{Knapsack}
\mvol{Don't say ``to close this section'' if it is not the last sub-section!}
To close this section, we show how our framework can represent more advanced
recursion schemes, such as \emph{dynamorphisms} for dynamic programming, by
using their encoding as a hylomorphism. We use the knapsack example
from~\cite{HinzeWG15}. First, we define memoisation tables in terms
of containers. 
\begin{coqcode}
Definition MemoShape : Type := A * Sg.
Definition MemoPos := Pg.
Instance Memo : Cont MemoShape MemoPos := { valid := valid \o pair (snd \o fst) snd }.

Definition Table := LFix Memo.
\end{coqcode}
Memoisation tables are the least fixed point of the container defined by shapes
\coq{A * Sg} and positions \coq{Pg}, given a a container \coq{G : Cont Sg Pg}.
We define a function to insert elements into the memoisation tables:
\begin{coqcode}
  Definition Cons : A * App G Table ~> App Memo Table := (* *)
\end{coqcode}
And two functions to inspect the head of a memoisation table, and remove an
element from it:
\begin{coqcode}
Definition headT : Table ~> A := (* *)
Definition tailT : Table ~> App G Table := (* *)
\end{coqcode}
Using these definitions, a dynamorphism is defined as follows:
\begin{coqcode}
Definition dyna (a : App G Table ~> A) (c : RCoalg G B) : B ~> A
:= headT \o hylo (l_in \o Cons \o pair a id) c.
\end{coqcode}
Note how, instead of an algebra \coq{App G A ~> A}, the algebra takes a
memoisation table. The definition of the algebra can use this table to lookup
elements, instead of triggering a further recursive call. Elements are inserted
into the memoisation table by the use of \coq{Cons} to the result of applying
the algebra. The algebra for the dynamorphism looks up the previously computed
elements to produce the result, thus saving the corresponding recursive calls:
\begin{coqcode}
Fixpoint memo_knap table wvs :=
  match wvs with | nil => nil | h :: t =>
      match lookupT (Nat.pred (fst h)) table with
      | Some u => (u + snd h)%sint63 :: memo_knapsack table t
      | None => memo_knapsack table t
      end
  end.

Definition knapsack_alg (wvs : list (nat * int))
  (x : App NatF (Table NatF int)) : int :=
  match x with | MkCont sx kx => match sx with
  | inl tt => fun kx => 0%sint63
  | inr tt => fun kx => let tbl := kx next_elem in max_int 0 (memo_knap tbl wvs)
  end kx end.
Definition knapsackA wvs : App NatF (Table NatF int) ~> int := 
  (* [knapsack_alg wvs] as a respecful morphism *)
\end{coqcode}
The hylomorphism for knapsack is as follows, where \coq{out_nat} is the
recursive coalgebra for \coq{nat}.
\begin{coqcode}
Example knapsack wvs : Ext (dyna (knapsackA wvs) out_nat).
\end{coqcode}
Coq's code extraction mechanism is unable to inline several definitions in
this case, but the extracted code can be trivially inlined to produce the
following:
\begin{ocamlcode}
let knapsack wvs x =
  let (y, _) =
    (let rec f x0 =
      if x0=0 then Uint63.of_int (0)
      else let fn := f (x0-1) in { lFix_out = { 
           shape = (max_int (Uint63.of_int (0)) (memo_knap fn wvs), sx);
           cont = fun _ -> fn } }
     in f x).lFix_out.shape
  in y
\end{ocamlcode}
Note how the recursive calls of \ocaml{f} build the memoisation table, and how
this memoisation table is used to compute the intermediate results in
\ocaml{memo_knap}, which is finally discarded to produce the final result.

\subsection{Shortcut Deforestation}
\mvol{Again, this needs a glue sentence, like: ``The final case study we consider is shortcut deforestation on lists.''}
Shortcut deforestation can be expressed succintly in terms of hylomorphisms and
their laws~\cite{TakanoM95}. In particular, given a function:
\begin{coqcode}
  s  : forall A. (App F A -> A) -> (App F A -> A)
\end{coqcode}
We can conclude, by parametricity, that
\begin{coqcode}
  hylo a l_out \o hylo (sigma l_in) c =e hylo (s a) c
\end{coqcode}
This is generally known as the \emph{acid rain} theorem. Unfortunately, this is
\textbf{not} provable in Coq if we want to remain axiom-free, since we would
need to add the necessary parametricity
axiom~\cite{keller_et_al:LIPIcs.CSL.2012.381}.

>>>>

Both mergesort and quicksort are divide-and-conquer algorithms that can be
captured by the structure of an hylomorphisms. The structure of the recursion
is that of a binary tree.  For example, in the case of quicksort, a list is
split into a pivot, the label of the node, and two sublists. We define the data
functor of trees as follows:
\begin{coqcode}
Inductive ITreeF L N X := i_leaf (l : L) | i_node (n : N) (l r : X)
\end{coqcode}
We define the functor as a container using the following shapes and positions:
\begin{coqcode}
Inductive Tshape L A := | Leaf (ELEM : L) | Node (ELEM : A).
Inductive Tpos := | Lbranch | Rbranch. 
\end{coqcode}
These define a container, \coq{TreeF}, in a straightforward way, by making
the positions of type \coq{Tpos} only valid in \coq{Node}.
We define a series of definitions for tree container constructors and 
destructors:
\mpav{The following algebra should be called a\_in or l\_in whatever}
\begin{coqcode}
Definition a_out {L A X : Type} : App (TreeF L A) X ~> ITreeF L A X.

@@

Definition c_split : Coalg (TreeF unit int) (list int).
|{ x ~> match x with
        | nil => a_leaf tt
        | cons h t => let (l, r) := List.partition (fun x => x <=? h) t in
                      a_node h l r
        end}|.
Defined.
\end{coqcode}
We prove that the coalgebra \coq{c_split} is recursive by showing
that it respects the ``less-than'' relation on the length of the lists.
%\begin{coqcode}
%Lemma split_fin : respects_relation c_split (@length int) lt.  (* *) Qed.
%
%Definition tsplit : RCoalg (TreeF (list int) unit) (list int)
%  := mk_wf_coalg wf_lt split_fin.
%\end{coqcode}
The code that we extract for \coq{hylo merge c_split} is the following:
\begin{ocamlcode}
let rec qsort = function
| [] -> [] | h :: t ->
  let (l, r) = partition (fun x0 -> leb x0 h) t in
  let x0 = fun e -> qsort (match e with | Lbranch -> l | Rbranch -> r) in
  app (x0 Lbranch) (h :: (x0 Rbranch))
\end{ocamlcode}
Note that Coq's code extraction is unable to inline \coq{x0}, but the resulting
code is similar to a hand-written \coq{qsort}.
The mergesort algorithm can be defined analogously and can be found in the
formalisation. % [REF LINK].

@@

% Putting everything together, Mergesort is defined as the following
% hylomorphism:
% \begin{coqcode}
% Definition msort := hylo merge tsplit.
% \end{coqcode}
% Note, however, that we could have composed the catamorphism and anamorphism,
% and used the deforestation result to optimise the intermediate binary tree
% away:
% \begin{coqcode}
% Definition msort : Ext (cata merge \o rana tsplit).
% calculate. rewrite (deforest l_in_out). simpl. reflexivity. Defined.
% \end{coqcode}
% In both cases, the result is the same, and it extracts to the usual divide and
% conquer definition of Mergesort in OCaml:
% \begin{ocamlcode}
% let rec msort x = match x with
% | [] -> x | _ :: l -> (match l with | [] -> x | _ :: _ ->
%      let (l0, r) = splitL x [] [] in
%      let cx = fun p -> match p with | Lbranch -> l0 | Rbranch -> r in
%      mergeL (msort (cx Lbranch)) (msort (cx Rbranch)))
% \end{ocamlcode}
\subsubsection{Fusing a divide-and-conquer computation}
As an example of how can we use program calculation techniques in our 
framework, we show how another traversal can be fused into the
divide-and-conquer algorithm using the laws of hylomorphisms.
Suppose that we map a function to the result of sorting the list. We can
use our framework to fuse both computations. In particular, consider
the following definition:
\begin{coqcode}
Definition qsort_times_two := Lmap times_two \o hylo merge tsplit.
\end{coqcode}
Here, \coq{Lmap times_two} is a list map function defined as a hylomorphism,
and \coq{times_two} multiplies every element of the list by two.
We can use Coq's generalised rewriting, and \coq{hylo_fusion_l} to fuse
\coq{times_two} into the RHS hylomorphism in \coq{qsort_times_two}.
After applying hylo fusion and the necessary rewrites, the hylomorphism that we
extract is \coq{hylo (merge \o natural times_two) tsplit}. In this definition,
\coq{natural} defines a natural transformation by applying \coq{times_two} to
the shapes, and \coq{times_two} multiplies every pivot in the Quicksort tree by
two. Our formalisation contains a proof that \coq{natural} is indeed a natural
transformation, which relies on the fact that it preserves the structure of the
shapes and, therefore, the validity of the positions.  The extracted OCaml code
is a single recursive traversal:
\begin{ocamlcode}
let rec qsort_times_two = function
| [] -> []
| h :: t ->
  let (l, r) = partition (fun x0 -> leb x0 h) t in
  let x0 = fun p -> qsort_times_two (match p with
                                     | Lbranch -> l
                                     | Rbranch -> r) in
  app (x0 Lbranch) ((mul (Uint63.of_int (2)) h) :: (x0 Rbranch))
\end{ocamlcode}

\subsection{Knapsack}
We focus now on the formalisation and extraction of \emph{dynamorphisms} for
dynamic programming, by using their encoding as a hylomorphism. We use the
knapsack example from~\cite{HinzeWG15}. Dynamorphisms build a memoisation table
that stores intermediate results, alongside the current computation. The
algebra used to define a dynamorphism can access this memoisation table to
speed up computation.  First, we define memoisation tables in terms of
containers. 
\begin{coqcode}
Definition MemoShape : Type := A * Sg.
Definition MemoPos := Pg.
Instance Memo : Cont MemoShape MemoPos := { valid := valid \o pair (snd \o fst) snd }.

Definition Table := LFix Memo.
\end{coqcode}
Memoisation tables are the least fixed point of the container defined by shapes
\coq{A * Sg} and positions \coq{Pg}, given a a container \coq{G : Cont Sg Pg}.
We define a function to insert elements into the memoisation tables:
\begin{coqcode}
  Definition Cons : A * App G Table ~> App Memo Table := (* *)
\end{coqcode}
And two functions to inspect the head of a memoisation table, and remove an
element from it:
\begin{coqcode}
Definition headT : Table ~> A := (* *)
Definition tailT : Table ~> App G Table := (* *)
\end{coqcode}
These tables map ``paths'' in the least fixed point of \coq{Memo} to elements
of type \coq{A}. For example, if \coq{G} is a list-generating functor, these
paths will be natural numbers. 
Using these definitions, a dynamorphism is defined as follows:
\begin{coqcode}
Definition dyna (a : App G Table ~> A) (c : RCoalg G B) : B ~> A
:= headT \o hylo (l_in \o Cons \o pair a id) c.
\end{coqcode}
Note how, instead of an algebra \coq{App G A ~> A}, the algebra takes a
memoisation table. The definition of the algebra can use this table to lookup
elements, instead of triggering a further recursive call. Elements are inserted
into the memoisation table by the use of \coq{Cons} to the result of applying
the algebra. The algebra for the dynamorphism looks up the previously computed
elements to produce the result, thus saving the corresponding recursive calls:
\begin{coqcode}
Fixpoint memo_knap table wvs :=
  match wvs with | nil => nil | h :: t =>
      match lookupT (Nat.pred (fst h)) table with
      | Some u => (u + snd h)%sint63 :: memo_knapsack table t
      | None => memo_knapsack table t
      end
  end.

Definition knapsack_alg (wvs : list (nat * int))
  (x : App NatF (Table NatF int)) : int :=
  match x with | MkCont sx kx => match sx with
  | inl tt => fun kx => 0%sint63
  | inr tt => fun kx => let tbl := kx next_elem in max_int 0 (memo_knap tbl wvs)
  end kx end.
Definition knapsackA wvs : App NatF (Table NatF int) ~> int := 
  (* [knapsack_alg wvs] as a respecful morphism *)
\end{coqcode}
The hylomorphism for knapsack is as follows, where \coq{out_nat} is the
recursive coalgebra for \coq{nat}.
\begin{coqcode}
Example knapsack wvs : Ext (dyna (knapsackA wvs) out_nat).
\end{coqcode}
Coq's code extraction mechanism is unable to inline several definitions in this
case. We have manually inlined the extracted code for simplicity. The reader
can check in our \textbf{Artefact} that the extracted code can be trivially
inlined to produce the following:
\begin{ocamlcode}
let knapsack wvs x =
  let (y, _) =
    (let rec f x0 =
      if x0=0 then Uint63.of_int (0)
      else let fn := f (x0-1) in { lFix_out = { 
           shape = (max_int (Uint63.of_int (0)) (memo_knap fn wvs), sx);
           cont = fun _ -> fn } }
     in f x).lFix_out.shape
  in y
\end{ocamlcode}
Note how the recursive calls of \ocaml{f} build the memoisation table, and how
this memoisation table is used to compute the intermediate results in
\ocaml{memo_knap}, which is finally discarded to produce the final result.

\subsection{Shortcut Deforestation}
The final case study we consider is shortcut deforestation on lists.  Shortcut
deforestation can be expressed succintly in terms of hylomorphisms and their
laws~\cite{TakanoM95}. In particular, given a function:
\begin{coqcode}
  s  : forall A. (App F A -> A) -> (App F A -> A)
\end{coqcode}
We can conclude, by parametricity, that
\begin{coqcode}
  hylo a l_out \o hylo (sigma l_in) c =e hylo (s a) c
\end{coqcode}
This is generally known as the \emph{acid rain} theorem. Unfortunately, this is
\textbf{not} provable in Coq if we want to remain axiom-free, since we would
need to add the necessary parametricity
axiom~\cite{keller_et_al:LIPIcs.CSL.2012.381}.
